### 2024-06-01 Manuscript/main.tex
main_PC.tex

\noindent RH: The what how and why of trait-based analyses

@@

\section{Abstract}
Objective: Showing which patterns are recovered by different functional diversity metrics under different frameworks and how different metrics recover different signals from different processes linked to common mechanisms in ecology.

Approach: We simulate some one dimensional datasets using neutral evolutionary models (pure birth tree with brownian motion traits) then perturb the resulting dataset with some common ecological stressor mechanisms (environmental filtering, competitive exclusion, equalising fitness and facilitation).
We then measure different space occupancy metrics on the perturbed and unperturbed simulations and check how the different metrics capture the resultant changes.

@@

In the last two decades, there has been a progressive shift in ecology and evolution from  taxonomic-oriented approaches, with species as the focal point of studies, to functional approaches that place species-specific characteristics (traits) at the centre of analyses \citep{mammola2021concepts, palacio2022protocol}.
Although numerous definitions exist \citep{dawson2021traits}, we here consider a trait to be any observable characteristic (e.g. morphological, anatomical, ecological, physiological, behavioural, phenological, etc.) measured on individual organisms at any level, from genes to whole organisms.
A trait-based approach has two advantages for answering core questions in ecology and evolution.
First, it allows a deeper understanding of the mechanisms generating biodiversity patterns, by putting organisms' traits at the centre of natural selection rather than organisms themselves.
Second, it allows comparisons across subdisciplines in biology, while facilitating the conceptualization of general principles broadly valid in space (e.g. unrelated species pools) and time (e.g. anatomical traits are comparable between palaeontology and ecology, which is not always the case with species; \citealt{luza2023going}).
Already used routinely in palaeontology \citep{raup1961geometry, gould1991disparity, foote1995morphological, guillerme2020disparities}, this new trait-focused ecology and evolution is unlocking the possibility to answer a broad range of questions in disciplines as diverse as community ecology \citep{mcgill2006rebuilding}, biogeography \citep{violle2014emergence}, conservation biology \citep{chichorro2022trait}, micro- \citep{chapin1993evolution}, macro-evolution \citep{guillerme2023innovation} and applied fields (e.g. agronomy \citealt{martin2015plant}).
This is because trait-based approaches closely align with the general analytical framework proposed by \cite{anand1994pattern} to answer three consequential questions: \textit{``what?''} (describing the pattern), \textit{``how?''} (describing the process) and \textit{``why?''} (understanding the mechanism) (see Box 1).

@@

\section*{Box 1}
\cite{anand1994pattern} conceptualised a general analytical framework for answering scientific questions in ecology and evolution, based on three consequential questions: \textit{``what?''} (describing the pattern), \textit{``how?''} (describing the process) and \textit{``why?''} (understanding the mechanism).
This framework can be effectively applied to trait-based analyses:

\textit{What?} Pattern description corresponds to the steps needed to collect and summarise data to answer a question of interest (i.e., ``how?'' and ``why?'' as defined below).
Usually this consists in: i) collecting target traits at the focal level (e.g., genes, individual, population, species; \citealt{violle2007let}); ii) arranging these traits into some kind of trait space \citep{guillerme2020disparities, mammola2021concepts}; and iii) using one or more unidimensional statistics to summarise properties of the trait space.
For unidimensional spaces (i.e. distributions), these statistics can be as simple as the mean and the standard deviation (e.g. community weighted mean).
For multidimensional spaces, these are usually named disparity metrics or indices \citep{guillerme2020disparities} or functional diversity metrics \citep{mammola2021concepts}, but all attempt to capture some pattern of interest in the trait space.

@@

While the first part of this framework (``what?'') has been studied extensively in the last decade this was often undertaken outside of the ``what, how and why'' framework.
Here we argue that the pattern description, although the basis of any future study, often follows some framework without a proper evaluation of its adequacy to answer the how and why.
The choice of the tool or metric used to describe the pattern is crucial for allowing us to understand the process and the mechanism.
Using an inappropriate metric for describing a pattern can lead to biased conclusions.
For example, let us imagine one is interested in understanding how two populations compete with each other for a resource (the mechanism - ``why?'') on two islands, one home to both populations and another one home to only one of the population (the process - ``how?'').
In such a scenario, measuring the occupied area (i.e. $km^{2}$) of each population on each island (a pattern - ``what?'') will not be the most appropriate way to understand the potential competition between these populations.
In this simplistic example, the functional overlap between populations (``what and how'') might be more appropriate.
See for example \citealt{carvalho2020decomposing} for a discussion of how Darwin's finches share resources depending on the existence or not of competition.

@@

    \end{tabular}
    \caption{\scriptsize{\textbf{Description of the stressors applied to the simulated data and the mechanism each approximated.} Here we distinguish between the biological mechanism we are trying to simulate and the algorithmic one we used to simulate it. We also provide a more detailed description of the stressor. Note that in biological data, we don't expect any of the mechanisms to be applied to the data alone (e.g. equalising fitness and facilitation can both act on the distribution of species traits). Nor do we expect their effects to be unidirectional (for example, equalising fitness can happen both by removing the edges or changing the position of a trait distribution). These mechanisms serve as a simplified description of reality for the narrative purpose of this paper.}
}
    \label{Tab:mechanisms}
\end{table}

@@

\textbf{Aspect} & \textbf{Method} & \textbf{Name} & \textbf{Description} & \textbf{\texttt{R} function} & \textbf{Reference}\\
\hline
 & Trait Probability Density & Functional evenness & An index of the spread of the trait distribution in an even spread of the distribution (e.g. the height of bins in a histogram) & \texttt{TPD::REND} & \cite{carmona2019trait}\\
 & \cr 
 & Kernel density hypervolume & Functional evenness &  Overlap between a hypervolume and an imaginary hypervolume where traits are evenly distributed within their possible range & \texttt{BAT::kernel.evenness} & \cite{mammola2020functional}\\
 & \cr 
 & Dendrogram & Functional evenness & Regularity of abundances and distances between observations in the tree. & \texttt{BAT::evenness} & \cite{cardoso2015bat,cardoso2024calculating}\\
 & \cr 
 & Distances & Functional evenness & Regularity of abundances and distances along the minimum spanning tree linking all observations & \texttt{FD::dbFD()\$FEve} & \cite{Laliberte2010FD,villeger2008new}\\
\hline
Divergence & Trait Probability Density & Divergence & Overlap between the TPD function of a single observation and the TPD function of the whole set of observations & \texttt{TPD::REND} & \cite{carmona2019trait}\\
 & \cr 
 & Kernel density hypervolume & Functional dispersion  & Average distance between an observation and a sample of random points within and the probabilistic hypervolume & \texttt{BAT::kernel.dispersion} & \cite{mammola2020functional}\\
 & \cr 
 & Dendrogram & Functional dispersion & Average dissimilarity between any two observations in the dendrogram & \texttt{BAT::dispersion} & \cite{cardoso2015bat}\\
 & \cr  
 & Distances & Rao's quadratic entropy & The mean Euclidean distance between every two random pairs of species & \texttt{FD::dbFD()\$RaoQ} & \cite{botta2005rao}\\
\hline
Richness & Trait Probability Density & Functional richness & Sum of cells where trait probability density $> 0$ & \texttt{TPD::REND} & \cite{carmona2019trait}\\
 & \cr 
 & Kernel density hypervolume & Functional richness & Size (volume) of the hypervolume & \texttt{BAT::kernel.alpha} & \cite{mammola2020functional}\\
 & \cr
 & Convex Hull & Functional richness & Volume of the convex hull & \texttt{BAT::hull.alpha} & \cite{cornwell2006trait}\\
 & \cr
 & Dendrogram & Alpha diversity & Total branch length of the functional dendrogram & \texttt{BAT::alpha} & \cite{petchey2002functional,cardoso2015bat}\\
\hline

@@

The full trait dataset comprised 118 bird species, all native species known to have existed on these islands over the last 125,000 years.
55 species are known to have gone extinct prior to 1500 CE and 26 after 1500 CE.
For our analyses here, we focused on two time periods: (i) the avifauna present in 1500 CE (the ``historic dataset'') and (ii) the current native avifauna (``extant dataset'').
We represented the distinct reductions in species richness preceding these two time periods as our two stressors, the first one representing all pre-1500 CE extinctions (i.e., $118-55$ species; a species richness reduction of 46\%) and the second all the extinctions that have occurred until the present (i.e., $118-(55+26)$ species; a species richness reduction of 69\%).

@@

We tested 12 trait space occupancy metrics on simulated and empirical datasets to assess how each metric captures patterns of trait space changes based on stressors approximating ecological and evolutionary mechanisms.
Our results show that, unexpectedly, different metrics capture different patterns (what) leading to inferring different processes (how) that can be proposed to explain different mechanisms (why).
Therefore, the choice of traits space occupancy metric is essential to accurately describing a pattern of interest.

Our results may be influenced by our choice of space occupancy metrics - specifically, we focused on metrics available in three statistical packages in \texttt{R}, the most common statistical language currently used in ecology \citep{lai2019evaluating}, but many other ones could have been used (e.g. \citealt{guillerme2020shifting}).
Also, our results are likely affected by the use of simplified algorithms designed to approximate very complex ecological and evolutionary mechanisms by simply removing a percentage of elements in trait spaces in a non-random manner.
Note also that the simulated trait spaces here did not always share common characteristics with empirical trait spaces.
The simulated trait spaces in 2, 4 or 8 dimensions had the same variance on all dimensions (i.e. we used the same trait simulation process for all dimensions) whereas in empirical trait spaces, commonly generated using ordination techniques (e.g. PCA or PCO), the dimensions have by definition a decreasing variance.

@@

Following our results and these thoughts, we would like to encourage workers in ecology and evolution to test whether their metrics are capable of capturing a pattern of interest before discussing the process and/or mechanism inferred by it.
We encourage workers to test their metrics using simplified example datasets to see if it has the ability to capture at least crude changes as exemplified in this work (e.g. using simulations \citealt{guillerme2024treats} or the \texttt{moms} interactive package \citealt{guillerme2020shifting}).
Importantly, it should always be borne in mind that empirical mechanisms ought of course to be more complex in real world scenarios.
For example, evolutionary mechanisms can vary through time or per clades (and so patterns are often the results of multiple processes); or ecological mechanisms are often intertwined and work together to generate a pattern (e.g. facilitation + competition; \citealt{danet2024species}), or counteract each others by operating on the trait space in opposite directions (e.g., competition + filtering; \citealt{mammola2024functional})

>>>>

\noindent RH: The what, how and why of trait-based analyses

@@

\section{Abstract}

Functional diversity is commonly used to describe populations and communities in ecology beyond taxonomic diversity.
It allows summarising complex space  or time-occupancy patterns (what is changing?) that can be used to understand changes in communities and/or populations (the process; how is it changing?) in response to some stressors (the mechanism; why is it changing?). 
Although researchers have the choice between many different metrics (what) to study their questions of interest (how and why), it is often not directly clear which metric is more readily appropriate for which question.
Here we show which patterns are recovered by multiple functional diversity metrics under different frameworks and how different metrics recover contrasting signals from different processes linked to common mechanisms in ecology.
Using both simulated data with common ecological mechanism approximations (environmental filtering, competitive exclusion, equalising fitness and facilitation) and an empirical dataset affected by more complex and nuanced mechanisms, we test the effectiveness of different space occupancy metrics to recover the simulated or empirical changes.
We show that metrics perform differently for many tasks and discuss the importance of not using a one-fit-all metric but rather think and test whether a particular one will be good at capturing a pattern of interest.

@@

In the last two decades, there has been a progressive expansion in ecology and evolution studies from taxonomic-oriented approaches, with species as the focal point, to functional approaches that place species-specific characteristics (traits) at the centre of analyses \citep{mammola2021concepts, palacio2022protocol}. Although numerous definitions exist \citep{dawson2021traits}, we here consider a trait to be any observable characteristic (e.g. morphological, anatomical, ecological, physiological, behavioural, phenological, etc.) measured on individual organisms at any level, from genes to whole organisms.
A trait-based approach has two advantages for answering core questions in ecology and evolution. First, it allows a deeper understanding of the mechanisms generating biodiversity patterns, by putting organisms' traits at the centre of natural selection rather than the organisms themselves. Second, it allows comparisons across subdisciplines in biology, while facilitating the conceptualization of general principles broadly valid in space (e.g. unrelated species pools) and time (e.g. anatomical traits are comparable between palaeontology and ecology, which is not always the case with species; \citealt{luza2023going}). Already used routinely in palaeontology \citep{raup1961geometry, gould1991disparity, foote1995morphological, guillerme2020disparities}, this new trait-focused ecology and evolution is unlocking the possibility to answer a broad range of questions in disciplines as diverse as community ecology \citep{mcgill2006rebuilding}, biogeography \citep{violle2014emergence}, conservation biology \citep{chichorro2022trait}, micro- \citep{chapin1993evolution}, macro-evolution \citep{guillerme2023innovation} and applied fields (e.g. agronomy \citealt{martin2015plant}). This is because trait-based approaches closely align with the general analytical framework proposed by \cite{anand1994pattern} to answer three sequential questions: \textit{``what?''} (describing the pattern), \textit{``how?''} (describing the process) and \textit{``why?''} (understanding the mechanism) (see Box 1).

@@

\section*{Box 1}
\cite{anand1994pattern} conceptualised a general analytical framework for answering scientific questions in ecology and evolution, based on three sequential questions: \textit{``what?''} (describing the pattern), \textit{``how?''} (describing the process) and \textit{``why?''} (understanding the mechanism).
This framework can be effectively applied to trait-based analyses:

\textit{What?} Pattern description corresponds to the steps needed to collect and summarise data to answer a question of interest (i.e., ``how?'' and ``why?'' as defined below).
Usually this consists in: i) collecting target traits at the focal level (e.g., genes, individual, population, species; \citealt{violle2007let}); ii) arranging these traits into some kind of trait space \citep{guillerme2020disparities, mammola2021concepts}; and iii) using one or more unidimensional or multidimensional statistics to summarise properties of the trait space. For unidimensional spaces (i.e. distributions), these statistics can be as simple as the mean and the standard deviation (e.g. community weighted mean). For multidimensional spaces, these are usually named disparity metrics or indices \citep{guillerme2020disparities} or functional diversity metrics \citep{mammola2021concepts}, but all attempt to capture some pattern of interest in the trait space.

@@

While the first part of this framework (``what?'') has been studied extensively in the last decade, this was often undertaken outside of the ``what, how and why'' framework. Here we argue that the pattern description, although the basis of any future study, often follows some framework without a proper evaluation of its adequacy to answer the how and why. The choice of the tool or metric used to describe the pattern is crucial for allowing us to understand the process and the mechanism. Using an inappropriate metric for describing a pattern can lead to biased conclusions.
For example, let us imagine one is interested in understanding how two populations compete with each other for a resource (the mechanism - ``why?'') on two islands, one home to both populations and another one home to only one of the population (the process - ``how?''). In such a scenario, measuring the occupied area (i.e. $km^{2}$) of each population on each island (a pattern - ``what?'') will not be the most appropriate way to understand the potential competition between these populations. In this simplistic example, the functional overlap between populations (``what and how'') might be more appropriate. See for example \citealt{carvalho2020decomposing} for a discussion of how Darwin's finches share resources depending on the existence or not of competition.

@@

    \end{tabular}
    \caption{\scriptsize{\textbf{Description of the stressors applied to the simulated data and the mechanism each approximated.} Here we distinguish between the biological mechanism we are trying to simulate and the algorithmic one we used to simulate it. We also provide a more detailed description of the stressor. Note that in biological data, we don't expect any of the mechanisms to act on communities alone (e.g. equalising fitness and facilitation can both act on the distribution of species traits). Nor do we expect their effects to be unidirectional (for example, equalising fitness can happen both by removing the edges or changing the position of a trait distribution). These mechanisms serve as a simplified description of reality for the narrative purpose of this paper.}
}
    \label{Tab:mechanisms}
\end{table}

@@

\textbf{Aspect} & \textbf{Method} & \textbf{Name} & \textbf{Description} & \textbf{\texttt{R} function} & \textbf{Reference}\\
\hline
Richness & Trait Probability Density & Functional richness & Sum of cells where trait probability density $> 0$ & \texttt{TPD::REND} & \cite{carmona2019trait}\\
 & \cr 
 & Kernel density hypervolume & Functional richness & Size (volume) of the hypervolume & \texttt{BAT::kernel.alpha} & \cite{mammola2020functional}\\
 & \cr
 & Convex Hull & Functional richness & Volume of the convex hull & \texttt{BAT::hull.alpha} & \cite{cornwell2006trait}\\
 & \cr
 & Dendrogram & Alpha diversity & Total branch length of the functional dendrogram & \texttt{BAT::alpha} & \cite{petchey2002functional,cardoso2015bat}\\
\hline
Divergence & Trait Probability Density & Divergence & Overlap between the TPD function of a single observation and the TPD function of the whole set of observations & \texttt{TPD::REND} & \cite{carmona2019trait}\\
 & \cr 
 & Kernel density hypervolume & Functional dispersion  & Average distance between an observation and a sample of random points within and the probabilistic hypervolume & \texttt{BAT::kernel.dispersion} & \cite{mammola2020functional}\\
 & \cr 
 & Dendrogram & Functional dispersion & Average dissimilarity between any two observations in the dendrogram & \texttt{BAT::dispersion} & \cite{cardoso2015bat}\\
 & \cr  
 & Distances & Rao's quadratic entropy & The mean Euclidean distance between every two random pairs of species & \texttt{FD::dbFD()\$RaoQ} & \cite{botta2005rao}\\
\hline
 & Trait Probability Density & Functional evenness & An index of the spread of the trait distribution in an even spread of the distribution (e.g. the height of bins in a histogram) & \texttt{TPD::REND} & \cite{carmona2019trait}\\
 & \cr 
 & Kernel density hypervolume & Functional evenness &  Overlap between a hypervolume and an imaginary hypervolume where traits are evenly distributed within their possible range & \texttt{BAT::kernel.evenness} & \cite{mammola2020functional}\\
 & \cr 
 & Dendrogram & Functional evenness & Regularity of abundances and distances between observations in the tree. & \texttt{BAT::evenness} & \cite{cardoso2015bat,cardoso2024calculating}\\
 & \cr 
 & Distances & Functional evenness & Regularity of abundances and distances along the minimum spanning tree linking all observations & \texttt{FD::dbFD()\$FEve} & \cite{Laliberte2010FD,villeger2008new}\\
\hline

@@

The full trait dataset comprised 118 bird species, all native species known to have existed on these islands over the last 125,000 years.
55 species are known to have gone extinct prior to 1500 CE and 26 after 1500 CE.
For our analyses, we focused on two time periods: (i) the avifauna present in 1500 CE (the ``historic dataset'') and (ii) the current native avifauna (``extant dataset'').
We represented the distinct reductions in species richness preceding these two time periods as our two stressors, the first one representing all pre-1500 CE extinctions (i.e., $118-55$ species; a species richness reduction of 46\%) and the second all the extinctions that have occurred until the present (i.e., $118-(55+26)$ species; a species richness reduction of 69\%).

@@

We tested 12 trait space occupancy metrics on simulated and empirical datasets to assess how each metric captures patterns of trait space changes based on stressors approximating ecological and evolutionary mechanisms.
Our results show that, unexpectedly, different metrics capture different patterns (what) leading to inferring different processes (how) that can be proposed to explain different mechanisms (why).
Therefore, the choice of trait space occupancy metric is essential to accurately describe a pattern of interest.

Our results may be influenced by our choice of space occupancy metrics - specifically, we focused on metrics available in three statistical packages in \texttt{R}, the most common statistical language currently used in ecology \citep{lai2019evaluating}, but many others could have been used (e.g. \citealt{guillerme2020shifting}).
Also, our results are likely affected by the use of simplified algorithms designed to approximate very complex ecological and evolutionary mechanisms by simply removing a percentage of elements in trait spaces in a non-random manner.
Note also that the simulated trait spaces here did not always share common characteristics with empirical trait spaces.
The simulated trait spaces in 2, 4 or 8 dimensions had the same variance on all dimensions (i.e. we used the same trait simulation process for all dimensions) whereas in empirical trait spaces, commonly generated using ordination techniques (e.g. PCA or PCO), the dimensions have by definition a decreasing variance.

@@

Following our results and these thoughts, we encourage workers in ecology and evolution to test whether their metrics are capable of capturing a pattern of interest before discussing the process and/or mechanism inferred by it.
We encourage workers to test their metrics using simplified example datasets to see if they have the ability to capture at least crude changes as exemplified in this work (e.g. using simulations \citealt{guillerme2024treats} or the \texttt{moms} interactive package \citealt{guillerme2020shifting}).
Importantly, it should always be borne in mind that empirical mechanisms ought of course to be more complex in real world scenarios.
For example, evolutionary mechanisms can vary through time or per clades (and so patterns are often the results of multiple processes); or ecological mechanisms are often intertwined and work together to generate a pattern (e.g. facilitation + competition; \citealt{danet2024species}), or counteract each others by operating on the trait space in opposite directions (e.g., competition + filtering; \citealt{mammola2024functional})
